NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-01 Manuscript/manuscriptPS.tex
Update text

Area per lipid is in agreement with experiments \cite{pan14} only
in the MacRog simulation with potassium counterions and in Gromos-CKP models,
while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, increasing counterion binding affinity reduces the area per molecule
when comparing between sodium and potassium in MacRog simulations or between
ff99 and jc ions in Lipid 17 simulations.
This suggests that the low area per molecule in PS lipid bilayers cannot be
explained only by counterion binding but also the hydrogen bonding network has to
be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Previously used method to evaluate counterion binding affinity to lipid bilayers
using the changes of hedgroup order parameters as a function
additional ions \cite{catte16} is difficult to apply for counterions because
reference state without ions in solution does not exist, see section \ref{mixtureTOadditionalCIs}
in supplementary information. However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
%The headgroup order parameters of POPC increase
%with increasing amount of negatively charged POPS
The experimentally measured order parameter increase~\cite{scherer87}
is reproduced in simulations with MacRog and potassium counterions, but not in 
other simulations (Fig. \ref{HGorderparametersPCvsPS}). 
The increase of POPC headgroup order parameters with increasing amount of POPS 
can be explained using the electrometer concept \cite{seelig87,scherer89}:
The PC headgroup tilts more parallel to the membrane plane with increasing
amount of negative charge in the membrane. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged PS
lipids and the order parameters actually decrease because also the amount of
positive counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Therefore, we conclude that the weak binding of potassium in the MacRog
simulations (Fig. \ref{NAdensPOPS}) leads to the correct surface charge
density and the response of POPC headgroup to the increasing amount of negatively
charges PS lipids in the lipid mixtures. 
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

>>>>

Area per lipid is in agreement with experiments \cite{pan14} only
in the MacRog simulation with potassium counterions and in Gromos-CKP models,
while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, increasing counterion binding affinity reduces the area per molecule
when comparing between sodium and potassium in MacRog simulations or between
ff99 and jc ions in Lipid 17 simulations.
This suggests that the counterion binding alone cannot explain the low area per molecule
in PS lipid bilayers but also the hydrogen bonding network has to
be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Evaluation of counterion binding affinity to charged lipid bilayers
using the changes of headgroup order parameters is difficult because
reference state without ions does not exist, in contrast to previous
studies with zwitterionic membranes and coions \cite{catte16},
see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}
